nextflu: Real-time tracking of seasonal influenza virus evolution in humans

### 2015-06-13 manuscript/correspondence/response.tex

The paper is interesting and clearly written so I don't have any major comment.

\subsection*{Minor revisions}

p1, col1, l47: ``near real-time'', can the authors specify the updating frequency? from the website, it seems to be every 5 days.

{\bf At the moment, GISAID does not provide an API to automatically download sequences, and so, as of now, we need to update manually, which involves clicking through to download all recent human sequences from GISAID and running the augur processing script. We monitor the addition of new sequences in the database and update the site when more than 10 new sequences have appeared relative to what's present on the nextflu site. In practice, this means updates every 3-7 days. We've made a note of this in the main text. From what we understand, a forthcoming version of the GISAID database will make the process of updating easier and allow automated downloads through an API. When this occurs, nextflu could be fully automated and update daily.}

@@

p1, col2, l54: how is the outgroup chosen?

{\bf We choose a well characterized strain predating the time interval of interest by a few years (A/Beijing/32/1992 for H3N2). We've updated the manuscript text to clarify this.}

p2, col1, l21: typo: the sampling time point is $t_v$ not $x(t_v)$

{\bf Fixed, thank you.}

@@

%%% REVIEWER 3 %%%
\section*{Reviewer 3}

\subsection*{Major comments}

This may be beyond the scope of the tool, but it would be extremely useful if the tool could visualize not just the \# of epitope or receptor binding site changes, but which specific amino acid changes are involved -- at least for the branches defining the main clusters identified (e.g., 3c3.a and 3c2.a).

{\bf This is a good suggestion. We've revised nextflu to make it so that tool tips of branches now show the amino-acid mutations associated with the branch. Furthermore, we've added a display of all alignment variability that allows one to spot variable positions. Clicking on these positions in the graph will plot the frequencies of amino acids and color the tree accordingly. This allows investigation of how specific amino acid changes map to the phylogeny.}

@@

\subsection*{Minor comment}

Could the authors clarify what they mean by `removing known reassortant clusters'?  This is very vague and I assume naturally occurring reassortant viruses are not being removed.

{\bf We filter sequences that are very similar to sequences from outbreaks of triple-reassortant swine influenza H1N1 that contain hemagglutinin (HA) from seasonal H3N2 viruses as described in Bastien et al. (2010, Journal of Infectious Diseases). This is a specific clade of reassortant viruses that has circulated sporadically in North America since 2009. We just filter out this clade, rather than looking for any reassortant virus. This has been clarified in the text.}

### 2015-06-13 manuscript/nextflu.tex
Further revise manuscript and response.

Every year, seasonal influenza infects between 10\% and 20\% of the global population resulting in substantial human morbidity and mortality \citep{flufactsheet}.
Vaccination remains the most effective public health measure to combat seasonal epidemics.
However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect those in the human population with acquired immunity to previously circulating strains.
Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data. By visualizing many different observables and predictors, we hope that \nextflu{} will help vaccine strain selection.
Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.

@@

In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
As input, \augur{} requires a FASTA file of sequences with FASTA header labels containing relevant information such as the strain name, the sampling date and passage history.
For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
The first step in the processing pipeline is to automatically select a subset of representative viruses.
Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
Following filtering, the viruses are subsampled to achieve a more equitable temporal and geographic distribution.
For our standard display period of 3 years and 50 viruses per month, this typically results in $\sim$1800 viruses, which we align using MAFFT \citep{katoh_mafft_2013}.
Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters. As outgroup for each viral lineage, we chose a well characterized virus without insertions relative to the canonical amino-acid numbering and a sampling date a few years before the time interval of interest.

@@

In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. Frequency are determined by maximizing the likelihood of sampling the observed set of virus sequences. In addition, we impose a smoothing that penalized rapid changes in frequency of the frequency derivative. \augur{} estimates frequency with up to one month resolution using a downhill simplex algorithm \citep{oliphant_python_2007}. At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.

These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and the phylogenetic tree displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
\begin{itemize}
	\item \textit{epitope mutations} at sites generally associated with antibody binding have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
%	\item non-epitope mutations: 
%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
    \item \textit{local branching index} is the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
 %   \item geographic location
    \item \textit{HA genotype} directly colors the tree by genotype at specific amino acid positions.
\end{itemize}
The tree display can also be restricted to different geographic regions.

@@

The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted. A second plot shows the variability of the alignment. On mouse-click on a variable positions in this plot, \auspice{} will color the tree by amino-acid at this position and plot the frequencies mutations at his positions.

@@

@@

\bibitem[Oliphant, 2007]{oliphant_python_2007}
Oliphant,T. (2007{\em{}}) Python for scientific computing.
\newblock {\em Comput Sci Eng, } {\bf 9} (3), 10--20.

>>>>

Every year, seasonal influenza infects between 10\% and 20\% of the global population resulting in substantial human morbidity and mortality \citep{flufactsheet}.
Vaccination remains the most effective public health measure to combat seasonal epidemics.
However, influenza viruses constantly evolve and thereby undergo antigenic drift, allowing drifted viruses to reinfect individuals with acquired immunity to previously circulating strains.
Owing to antigenic drift, the seasonal influenza vaccine needs frequent updating to remain effective.
In any given year, the particular choice of vaccine strain plays a major role in determining vaccine efficacy and so it is of critical importance to develop tools to analyze the ongoing evolution of the influenza virus population in order to aid vaccine strain selection.
The program \nextflu{} presents a near real-time display of genetic relationships among influenza viruses and allows investigation of currently available sequence data. By visualizing many different genetic and epidemiological features, we hope that \nextflu{} will help vaccine strain selection.
Currently, \nextflu{} tracks all four circulating lineages of seasonal influenza: A/H3N2, A/H1N1pdm, B/Victoria and B/Yamagata.

@@

In implementation, \nextflu{} consists of a processing pipeline written in Python called \augur{} that analyzes virus sequence data and a JavaScript-based browser visualization called \auspice{} that displays this processed information.
As input, \augur{} requires a FASTA file of sequences with header labels containing relevant information such as strain name, sampling date and passage history.
For this purpose, influenza sequence data for the hemagglutinin (HA) gene is downloaded from the GISAID EpiFlu database \citep{GISAID}, which contains the most up-to-date collection of seasonal influenza viruses.
The first step in the processing pipeline is to automatically select a subset of representative viruses.
Here, viruses without complete date or geographic information, viruses passaged in eggs and sequences less than 987 bases are removed.
In addition, local outbreaks are filtered by keeping only one instance of identical sequences sampled at the same location on the same day.
Following filtering, viruses are subsampled to achieve a more equitable temporal and geographic distribution.
For our standard display period of 3 years and 32 viruses per month, this typically results in $\sim$1200 viruses, for which we align full-length HA sequences where available and partial sequences otherwise, using MAFFT \citep{katoh_mafft_2013}.
Once aligned, the set of virus sequences is further cleaned by removing insertions relative to the outgroup to enforce canonical HA site numbering, by removing sequences that show either too much or too little divergence relative to the expectation given sampling date, and by removing known reassortant clusters, such as the triple-reassortant swine influenza viruses that have sporadically circulating since 2009 \citep{bastien_human_2010}. As outgroup for each viral lineage, we chose a well characterized virus without insertions relative to the canonical amino-acid numbering and a sampling date a few years before the time interval of interest.

@@

In addition to the phylogenetic tree, \augur{} estimates the frequency trajectories of mutations, genotypes and clades in the tree. Frequencies are determined by maximizing the likelihood of sampling the observed set of virus sequences. In addition, we impose a smoothing that penalized rapid changes in frequency of the frequency derivative. \augur{} estimates frequency with up to one month resolution. The result is similar to ``allele dynamics'' plots in \citet{steinbruck_allele_2011}, but provides frequencies of clades in the tree in addition to point mutations.
The \augur{} pipeline is run every 3-7 days in response to sequence updates in the GISAID database.

@@

At the end of the \augur{} pipeline, JSON files are exported containing the annotated phylogenetic tree, sequence data and frequency trajectories.
These JSON files are then visualized by \auspice{} using D$^3$ \citep{bostock_d3_2011} and a phylogenetic tree is displayed with branches scaled according to evolutionary distance across all sites (\FIG{tree}).
The user can explore the data interactively by selecting viruses from different dates or by coloring the tree by attributes such as:
\begin{itemize}
	\item \textit{epitope mutations} at sites generally associated with antibody binding that have been suggested to be predictive of future clade success \citep{luksza_predictive_2014},
%	\item non-epitope mutations: 
%	Mutations outside epitopes tend to be damaging and anti-correlated with clade expansion \citep{luksza_predictive_2014}.
    \item \textit{receptor binding mutations} at seven positions close to the receptor binding site that have been shown to be responsible for major antigenic transitions in the past decades \citep{koel_substitutions_2013},
    \item \textit{local branching index} indicating the exponentially weighted tree length surrounding a node, which is associated with rapid branching and expansion of clades \citep{neher_predicting_2014},
 %   \item geographic location
    \item \textit{HA genotype}, which directly colors the tree by genotype at specific amino acid positions.
\end{itemize}
The display can also be restricted to different geographic regions.

The frequency plot below the tree (\FIG{freq}) displays the frequency trajectory of clades in the tree whenever the mouse hovers above the branch defining the clade. Furthermore, trajectories of individual mutations, combinations of two mutations, and predefined clades such as 3c3.a can be plotted. A second plot shows the variability of the alignment. On mouse-click on a variable position in this plot, \auspice{} will color the tree by amino-acid at this position and plot its mutation frequencies.

@@

\bibitem[Bastien {\em et~al.}, 2010]{bastien_human_2010}
Bastien,N. {\em et~al.} (2010{\em{}})
  Human infection with a triple-reassortant swine influenza {A (H1N1)} virus
  containing the hemagglutinin and neuraminidase genes of seasonal influenza
  virus.
\newblock {\em J Infect Dis, } {\bf 201} (8), 1178--1182.

@@

%\bibitem[Oliphant, 2007]{oliphant_python_2007}
%Oliphant,T. (2007{\em{}}) Python for scientific computing.
%\newblock {\em Comput Sci Eng, } {\bf 9} (3), 10--20.
